Faithful Model Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-12-16 paper/camera-ready/altmeyer.tex
final things on main paper

Equation~\ref{eq:impl} gives rise to a similar evaluation metric for unfaithfulness. We swap out the subsample of observed individuals in the target class for the set of samples generated through SGLD ($\widehat{\mathbf{X}}_{\mathbf{y}^+}$):

@@

\begin{equation} \label{eq:eccco-start}
  \begin{aligned}
  \mathbf{Z}^\prime =& \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {L_{\text{JEM}}(f(\mathbf{Z}^\prime);M_{\theta},\mathbf{y}^+)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) } \} 
  \end{aligned} 
\end{equation}

@@

\begin{equation} \label{eq:eccco}
  \begin{aligned}
  \mathbf{Z}^\prime =& \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {L_{\text{clf}}(f(\mathbf{Z}^\prime);M_{\theta},\mathbf{y}^+)}+ \lambda_1 {\text{cost}(f(\mathbf{Z}^\prime)) } \\
  &+ \lambda_2 \mathcal{E}_{\theta}(f(\mathbf{Z}^\prime)|\mathbf{y}^+) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
  \end{aligned} 
\end{equation}

@@

To assess and benchmark the performance of our proposed generator against the state of the art, we generate multiple counterfactuals for different models and datasets. In particular, we compare \textit{ECCCo} and its variants to the following counterfactual generators that were introduced above: firstly; \textit{Schut}, which works under the premise of minimizing predictive uncertainty; secondly, \textit{REVISE}, which is state-of-the-art (SOTA) with respect to plausibility; and, finally, \textit{Wachter}, which serves as our baseline. In the case of \textit{ECCCo+}, we use principal component analysis (PCA) for dimensionality reduction: the latent space $\mathcal{Z}$ is spanned by the first $n_z$ principal components where we choose $n_z$ to be equal to the latent dimension of the VAE used by \textit{REVISE}.

For the predictive modelling tasks, we use multi-layer perceptrons (\textit{MLP}), deep ensembles, joint energy models (\textit{JEM}) and convolutional neural networks (LeNet-5 \textit{CNN}~\citep{lecun1998gradient}). Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of \textit{ECCCo}. To account for stochasticity, we generate multiple counterfactuals for each target class, generator, model and dataset.

We perform benchmarks on eight datasets from different domains. From the credit and finance domain we include three tabular datasets: Give Me Some Credit (\textit{GMSC})~\citep{kaggle2011give}, \textit{German Credit}~\citet{hoffman1994german} and \textit{California Housing}~\citet{pace1997sparse}. All of these are commonly used in the related literature~\citep{karimi2020survey,altmeyer2023endogenous,pawelczyk2021carla}. Following related literature~\citep{schut2021generating,dhurandhar2018explanations} we also include two image datasets: \textit{MNIST}~\citep{lecun1998mnist} and \textit{Fashion MNIST}~\citep{xiao2017fashion}.

Full details concerning model training as well as detailed descriptions and results for all datasets can be found in the appendix. In the following, we will focus on the most relevant results highlighted in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision}. The tables show sample averages along with standard deviations across multiple runs for our key evaluation metrics for the \textit{California Housing} and \textit{GMSC} datasets (Table~\ref{tab:results-tabular}) and the \textit{MNIST} dataset (Table~\ref{tab:results-vision}). For each metric, the best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). For the tabular datasets, we use the default Euclidian distance to measure unfaithfulness and implausibility as defined in Equations~\ref{eq:faith} and~\ref{eq:impl}, respectively. The third metric presented~\ref{tab:results-tabular} in Table quantifies the predictive uncertainty of the counterfactual as measured by Equation~\ref{eq:setsize}. For the vision datasets, we rely on measuring the structural dissimilarity between images for our unfaithfulness and implausibility metrics~\citep{wang2003multiscale}.

@@

Overall, we find strong empirical evidence suggesting that \textit{ECCCo} consistently achieves state-of-the-art faithfulness. Across all models and datasets highlighted here, all variations of \textit{ECCCo} consistently outperform other generators with respect to faithfulness, in many cases substantially. This pattern is mostly robust across all other datasets.

@@

While we maintain that generally speaking plausibility should hinge on the quality of the model, our results also indicate that it is possible to balance faithfulness and plausibility if needed: \textit{ECCCo+} generally outperforms other variants of \textit{ECCCo} in this context at the small cost of slightly reduced faithfulness. For the vision datasets especially, we find that  \textit{ECCCo+} is consistently second only to \textit{REVISE} for all models and regularly substantially better than the baseline. Looking at the \textit{California Housing} data, latent space search markedly improves plausibility without sacrificing faithfulness: for the \textit{JEM} Ensemble, \textit{ECCCo+} performs substantially better than the baseline and only marginally worse than \textit{REVISE}. Importantly, \textit{ECCCo+} does not attain plausibility at all costs: for the \textit{MLP} Ensemble, plausibility is still very low but this seems to faithfully represent what the model has learned.

>>>>

Equation~\ref{eq:impl} gives rise to a similar evaluation metric for unfaithfulness. We swap out the subsample of observed individuals in the target class for the set of samples generated through SGLD ($\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}$):

@@

\begin{equation} \label{eq:eccco-start}
  \begin{aligned}
  \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {L_{\text{JEM}}(f(\mathbf{Z}^\prime);M_{\theta},\mathbf{y}^+)}+ \lambda {\text{cost}(f(\mathbf{Z}^\prime)) } \} 
  \end{aligned} 
\end{equation}

@@

\begin{equation} \label{eq:eccco}
  \begin{aligned}
  & \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {L_{\text{clf}}(f(\mathbf{Z}^\prime);M_{\theta},\mathbf{y}^+)}+ \lambda_1 {\text{cost}(f(\mathbf{Z}^\prime)) } \\
  &+ \lambda_2 \mathcal{E}_{\theta}(f(\mathbf{Z}^\prime)|\mathbf{y}^+) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
  \end{aligned} 
\end{equation}

@@

To assess and benchmark the performance of our proposed generator against the state of the art, we generate multiple counterfactuals for different models and datasets. In particular, we compare \textit{ECCCo} and its variants to the following counterfactual generators that were introduced above: firstly, \textit{Schut}, which works under the premise of minimizing predictive uncertainty; secondly, \textit{REVISE}, which is state-of-the-art (SOTA) with respect to plausibility; and, finally, \textit{Wachter}, which serves as our baseline. In the case of \textit{ECCCo+}, we use principal component analysis (PCA) for dimensionality reduction: the latent space $\mathcal{Z}$ is spanned by the first $n_z$ principal components where we choose $n_z$ to be equal to the latent dimension of the VAE used by \textit{REVISE}.

For the predictive modelling tasks, we use multi-layer perceptrons (\textit{MLP}), deep ensembles, joint energy models (\textit{JEM}) and convolutional neural networks (LeNet-5 \textit{CNN}~\citep{lecun1998gradient}). Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of \textit{ECCCo}. To account for stochasticity, we generate many counterfactuals for each target class, generator, model and dataset over multiple runs.

We perform benchmarks on eight datasets from different domains. From the credit and finance domain we include three tabular datasets: Give Me Some Credit (\textit{GMSC})~\citep{kaggle2011give}, \textit{German Credit}~\citep{hoffman1994german} and \textit{California Housing}~\citep{pace1997sparse}. All of these are commonly used in the related literature~\citep{karimi2020survey,altmeyer2023endogenous,pawelczyk2021carla}. Following related literature~\citep{schut2021generating,dhurandhar2018explanations} we also include two image datasets: \textit{MNIST}~\citep{lecun1998mnist} and \textit{Fashion MNIST}~\citep{xiao2017fashion}.

Full details concerning model training as well as detailed descriptions and results for all datasets can be found in the appendix. In the following, we will focus on the most relevant results highlighted in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision}. The tables show sample averages along with standard deviations across multiple runs for our key evaluation metrics for the \textit{California Housing} and \textit{GMSC} datasets (Table~\ref{tab:results-tabular}) and the \textit{MNIST} dataset (Table~\ref{tab:results-vision}). For each metric, the best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). For the tabular datasets, we use the default Euclidian distance to measure unfaithfulness and implausibility as defined in Equations~\ref{eq:faith} and~\ref{eq:impl}, respectively. The third metric presented in Table~\ref{tab:results-tabular} quantifies the predictive uncertainty of the counterfactual as measured by Equation~\ref{eq:setsize}. For the vision datasets, we rely on measuring the structural dissimilarity between images for our unfaithfulness and implausibility metrics~\citep{wang2003multiscale}.

@@

Overall, we find strong empirical evidence suggesting that \textit{ECCCo} consistently achieves state-of-the-art faithfulness. Across all models and datasets highlighted here, different variations of \textit{ECCCo} consistently outperform other generators with respect to faithfulness, in many cases substantially. This pattern is mostly robust across all other datasets.

@@

While we maintain that generally speaking plausibility should hinge on the quality of the model, our results also indicate that it is possible to balance faithfulness and plausibility if needed: \textit{ECCCo+} generally outperforms other variants of \textit{ECCCo} in this context, occasionally at the small cost of slightly reduced faithfulness. For the vision datasets especially, we find that  \textit{ECCCo+} is consistently second only to \textit{REVISE} for all models and regularly substantially better than the baseline. Looking at the \textit{California Housing} data, latent space search markedly improves plausibility without sacrificing faithfulness: for the \textit{JEM} Ensemble, \textit{ECCCo+} performs substantially better than the baseline and only marginally worse than \textit{REVISE}. Importantly, \textit{ECCCo+} does not attain plausibility at all costs: for the \textit{MLP} Ensemble, plausibility is still very low but this seems to faithfully represent what the model has learned.
